Type-Safe Modular Parsing

### 2016-11-15 paper/sections/Library/DifferentSyntax.tex
Organize.
Some changes. Will reorganize some stuff in next commit.

\subsection{Object Algebras for Different Syntax}\label{subsec:differentsyntax}

Such extensibility becomes fragile when it comes to grammars with different syntax. Now suppose we would like to have types in the grammar. Firstly the new grammar is shown below, where the \lstinline{IntT} type and lambda expressions with annotation have been added:
\begin{lstlisting}
e ::= x | e ' ' e | \x:t.e
t ::= "Int"
\end{lstlisting}

@@

val pET : (=> Pair) => Pair = {
    val pE = pVar2 ||| pApp2 ||| pLam
    val pT = pInt
    p => (pE(p), pT(p))
}
\end{lstlisting}
Based on the old \lstinline{pVar} and \lstinline{pApp}, we implement \lstinline{pVar2} and \lstinline{pApp2} which take the pair as their fix-points instead.
In that case the parsers are consistent, and hence can be composed using the combinator \lstinline{|||}.
The last value \lstinline{pET} integrates the four single parsers defined before, where parsers for expressions and parsers for types are composed respectively. Furthermore, we use \lstinline{fix(p)} to obtain the pair of two parsers we want: one for expressions, the other for types. The parser \lstinline{fix(p)._1} parsers expressions like \lstinline{"\x:Int.x x"} successfully. Yet meanwhile, some drawbacks of this approach come into sight. With different syntax, it
is necessary to extend the fix-point to contain more parsers, in which case usually some structures like tuples are used,
making code tedious and less elegant. Moreover, we need to update such a fix-point in the arguments of all previously defined parsers, so as to make parsers consistent for the combinator (things get even worse without that combinator).
On the other hand, the modularity is quite restricted; only parsing and datatypes are extensible.
Suppose now we want to realize some behaviors on those data structures, for example pretty-printing, then whenever we introduce a new behavior,
it needs implementations in every class body, requiring existing code to be modified. At this point, Object Algebras are adopted for data structures
instead.

One intuition of using Object Algebras is to separate data structures from behaviors, so as to modularize the design, and both parts are even more modular and reusable. We can simply add new variants corresponding to the new cases in the grammar, or realize different operations on structures, without modifying existing code, meanwhile separate compilation is supported. Another one is that Object Algebras benefit us from using generics, where we can abstract different syntax in the type parameters, and define less classes, simplifying code comparatively. Below we present the code using Object Algebras, where we start from \lstinline{varE} and \lstinline{appE} only, later we will extend them with \lstinline{intT} and \lstinline{lamE}:
\begin{lstlisting}
trait OldAlg[E] {
    def varE(x : String) : E
    def appE(e1 : E, e2 : E) : E
}

trait OldParser[E] {
    val pE : OldAlg[E] => (=> Parser[E]) => Parser[E] = alg => p =>
        ident ^^ { x => alg.varE(x) } |||
        p ~ p ^^ { case e1 ~ e2 => alg.appE(e1, e2) }
}
\end{lstlisting}
Here \lstinline{OldAlg} is an Object Algebra interface. An Object Algebra interface is more like a small language, where several cases (constructors) are put together. Since with Object Algebras structures are implicit, a parser of \lstinline{OldAlg} should generate ``structures'' with
type \lstinline{"OldAlg[E] =>} \lstinline{E"} for any \lstinline{E}. In \lstinline{OldParser}, \lstinline{alg} of type
\lstinline{OldAlg[E]} can be any algebra, which appears as a parameter of \lstinline{pE}, then in the body of \lstinline{pE} the algebra
is invoked correspondingly right after parsing. It means we do not get any explicit objects generated from parsers, but they are immediately applied with algebras and return the results. To make \lstinline{alg} as a generic algebra, the enclosing trait of the parser, namely \lstinline{OldParser}, is generic.

Now we move on to declare \lstinline{intT} and \lstinline{lamE} in a new trait:
\begin{lstlisting}
trait NewAlg[E, T] {
    def intT() : T
    def lamE(x : String, t : T, e : E) : E
}

>>>>

\subsection{Parsing with Multiple Syntax}\label{subsec:differentsyntax}

\haoyuan{Another way to organize the paper is to introduce the lack support of class approach for multiple different syntax, and directly extend var and app with lam.}

Back to the class approach, the only little extensibility is just fragile when it comes to grammars with different syntax. Suppose we would like to have types in the expression language. Firstly the new grammar is shown below, where the \lstinline{Int} type and lambda expressions with annotation have been added to class \lstinline{Expr}:
\begin{lstlisting}
e ::= x | e ' ' e | \x:t.e
t ::= "Int"
\end{lstlisting}

@@

val pET : (=> Pair) => Pair = {
    val pE = pVar2 ||| pApp2 ||| pLam
    val pT = pInt
    p => (pE(p), pT(p))
}
\end{lstlisting}
Based on the old \lstinline{pVar} and \lstinline{pApp}, we implement \lstinline{pVar2} and \lstinline{pApp2} which take the pair as their fix-points instead.
In that case the parsers are consistent, and hence can be composed using the combinator \lstinline{|||}.
The last value \lstinline{pET} integrates the four single parsers defined before, where parsers for expressions and parsers for types are composed respectively. Furthermore, we use \lstinline{fix(p)} to obtain the pair of two parsers we want: one for expressions, the other for types. The parser \lstinline{fix(p)._1} parsers expressions like \lstinline{"\x:Int.x x"} successfully. Yet meanwhile, some drawbacks of this approach come into sight. With different syntax, it
is necessary to extend the fix-point to contain more parsers, in which case usually some structures like tuples are used,
making code tedious and less elegant. Moreover, we need to update such a fix-point in the arguments of all previously defined parsers, so as to make parsers consistent for the combinator (things get even worse without that combinator).

One intuition of using Object Algebras, as illustrated above, is to separate data structures from behaviors, so as to modularize the design, and both parts are even more modular and reusable. We can simply add new variants corresponding to the new cases in the grammar, or realize different operations on structures, without modifying existing code, meanwhile separate compilation is supported. Another one is that Object Algebras benefit us from using generics, where we can abstract different syntax in the type parameters, and define less classes, simplifying code comparatively. Below we present the code using Object Algebras, where we start from \lstinline{ExprAlg} (with \lstinline{varE} and \lstinline{appE} only), and define a new Object Algebra interface with \lstinline{intT} and \lstinline{lamE}:
\begin{lstlisting}
trait LamAlg[E, T] {
    def intT() : T
    def lamE(x : String, t : T, e : E) : E
}

### 2016-11-15 paper/sections/ObjectAlgebras.tex
Some changes. Will reorganize some stuff in next commit.

\subsection{Object Algebras}\label{sec:objectalgebras}
Object Algebras captures a design pattern to address the Expression Problem nicely,
achieving two dimensions of extensibility (data variants and operations) in a modular and type-safe way.
Because of this, the definition of data structures is separated from behaviors on them, and future extensions
to both sides no longer require existing code to be modified, supporting separate compilation.

In Object Algebras, ASTs as recursive data structures are defined using traits, where each constructor corresponds
to an abstract method inside. The recursive types are represented using generics. Here we show the example of a simple
expression language:
\begin{lstlisting}
trait ExpAlg[E] {
    def lit(i : Int) : E
    def add(e1 : E, e2 : E) : E
}
\end{lstlisting}
The language has two constructors: literals and additions. The trait \lstinline{ExpAlg} is called an \textit{Object Algebra interface},
whereas as a factory, it cannot produces objects for the expressions directly like traditional approaches, but abstract the results in its
type parameter \lstinline{E} instead. To realize an operation on expressions, we simply instantiate the type parameter by a concrete type and
provides implementations for all cases. Below is an example of evaluation:
\begin{lstlisting}
trait Eval extends ExpAlg[Int] {
    def lit(i : Int) = i
    def add(e1 : Int, e2 : Int) = e1 + e2
}
\end{lstlisting}
Here \lstinline{Eval} is called an \textit{Object Algebra}. It traverses an expression bottom-up, and returns an integer as the result.
Furthermore, one can define a new trait to implement pretty-printing on expressions:
\begin{lstlisting}
trait Print extends ExpAlg[String] {
    def lit(i : Int) = "" + i
    def add(e1 : String, e2 : String) = e1 + " + " + e2
}
\end{lstlisting}
On the other hand, the data variants can be extended by inheriting \lstinline{ExpAlg} and adding new cases only. Suppose we want to
have subtraction in expressions, a new Object Algebra interface \lstinline{SubAlg} is defined as follows:
\begin{lstlisting}
trait SubAlg[E] extends ExpAlg[E] {
    def sub(e1 : E, e2 : E) : E
}
\end{lstlisting}
Now evaluation on the new language can be realized by code reuse, and without modifying existing code:
\begin{lstlisting}
trait SubEval extends Eval {
    def sub(e1 : Int, e2 : Int) = e1 - e2
}
\end{lstlisting}
On this, to create an expression of \lstinline{SubAlg}, a generic method is defined as follows:
\begin{lstlisting}
def build[E](alg : SubAlg[E]) : E =
    alg.sub(alg.add(alg.lit(2), alg.lit(3)), alg.lit(4))
\end{lstlisting}
Such a method implicitly represents the expression \lstinline{"2 + 3 - 4"}. The code
\begin{lstlisting}
build(new Eval(){})
\end{lstlisting}
results in \lstinline{1}, as the result of evaluation. Now we observe that an ``object'' of expression actually
has the function type \lstinline{SubAlg[E] =>} \lstinline{E} for generic \lstinline{E}. At this point, one may argue
that it is illusory compared to traditional objects, and since we are struggling against the parsing problem, if a parser
produces structures as functions, it could be hardly be further processed like objects, especially, in the design of a compiler
there will be a lot of desugarings (or transformations). However, we argue that this is not an issue, because functions can be used
in a very clever way. For example, if one wants to refactor an \lstinline{ExpAlg} expression to a \lstinline{SubAlg} expression by
replacing all negative numbers with subtraction, a transformation algebra can be implemented as follows:
\begin{lstlisting}
trait Refactor[E] extends ExpAlg[E] {
    def alg : SubAlg[E]
    def lit(i : Int) = if (i > 0) alg.lit(i) else alg.sub(alg.lit(0), alg.lit(-i))
    def add(e1 : E, e2 : E) = alg.add(e1, e2)
}
\end{lstlisting}
With \lstinline{Refactor}, the expression \lstinline{"1 + (-2)"} is refactored to \lstinline{"1 + (0 - 2)"}. The abstract method
\lstinline{alg} in \lstinline{Refactor} can be any algebra of \lstinline{SubAlg}, simply for delegation. By passing a concrete implementation
to \lstinline{alg}, \lstinline{Refactor} will be able to apply refactoring to an \lstinline{ExpAlg} expression before \lstinline{alg} returns the result.

>>>>

\subsection{Object Algebras}\label{subsec:objectalgebras}
Object Algebras captures a design pattern to address the Expression Problem nicely,
achieving two dimensions of extensibility (data variants and operations) in a modular and type-safe way.
Because of this, the definition of data structures is separated from behaviors on them, and future extensions
to both sides no longer require existing code to be modified, supporting separate compilation.

In Object Algebras, ASTs as recursive data structures are defined using traits, where each constructor corresponds
to an abstract method inside. The example from Section~\ref{subsec:parsingwithopen} is again used here for illustration.
At first the language only supports variables and applications:
\begin{lstlisting}
trait ExprAlg[E] {
    def varE(x : String) : E
    def appE(e1 : E, e2 : E) : E
}
\end{lstlisting}
The language has two constructors: literals and additions. The trait \lstinline{ExprAlg} is called an \textit{Object Algebra interface},
whereas as a factory, it cannot produces objects for the expressions directly like traditional approaches, but abstract the results in its
type parameter \lstinline{E} instead. To realize an operation on expressions, we simply instantiate the type parameter by a concrete type and
provides implementations for all cases. Below is an example of collecting all free variables in an expression:
\begin{lstlisting}
trait FreeVars extends ExprAlg[List[String]] {
    def varE(x : String) = List(x)
    def appE(e1 : List[String], e2 : List[String]) = e1 ++ e2
}
\end{lstlisting}
Here \lstinline{FreeVars} is called an \textit{Object Algebra}. It traverses an expression bottom-up, and returns a list of strings as the result.
Furthermore, one can define a new trait to implement pretty-printing on expressions:
\begin{lstlisting}
trait ExprPrint extends ExprAlg[String] {
    def varE(x : String) = x
    def appE(e1 : String, e2 : String) = "(" + e1 + " " + e2 + ")"
}
\end{lstlisting}
On the other hand, the data variants can be extended by inheriting \lstinline{ExprAlg} and adding new cases only. Suppose we want to
have literals in expressions, a new Object Algebra interface \lstinline{LitAlg} is defined as follows:
\begin{lstlisting}
trait LitAlg[E] extends ExprAlg[E] {
    def litE(n : Int) : E
}
\end{lstlisting}
Now pretty-printing on the new language can be realized by code reuse, without modifying existing code:
\begin{lstlisting}
trait LitPrint extends ExprPrint {
    def litE(n : Int) = n.toString
}
\end{lstlisting}
On this, to create an expression of \lstinline{LitAlg}, a generic method is defined as follows:
\begin{lstlisting}
def build[E](alg : LitAlg[E]) : E =
    alg.appE(alg.varE("x"), alg.litE(3))
\end{lstlisting}
Such a method implicitly represents the expression \lstinline{"x 3"}. The code
\begin{lstlisting}
build(new LitPrint(){})
\end{lstlisting}
results in \lstinline{"(x 3)"}, as the result of pretty-printing. Now we observe that an ``object'' of expression actually
has the function type \lstinline{LitAlg[E] =>} \lstinline{E} for generic \lstinline{E}. At this point, one may argue
that it is illusory compared to traditional objects, and since we are struggling against the parsing problem, if a parser
produces structures as functions, it could be hardly be further processed like objects, especially, in the design of a compiler
there will be a lot of desugarings (or transformations). However, we argue that this is not an issue, because functions can be used
in a very clever way. For example, if one wants to transform an \lstinline{ExprAlg} expression to a \lstinline{LitAlg} expression by
replacing some variables with values based on a variable environment, a transformation algebra can be implemented as follows:
\begin{lstlisting}
trait Refactor[E] extends ExprAlg[E] {
    def alg : SubAlg[E]
    def env : Map[String, Int]
    def varE(x : String) = if (env.contains(x)) alg.litE(env(x)) else alg.varE(x)
}
\end{lstlisting}\haoyuan{Is this example OK?}
When \lstinline{env = Map("y"} \lstinline{-> 3)}, the \lstinline{ExprAlg} expression \lstinline{"x y"} is changed to \lstinline{LitAlg} expression \lstinline{"x 3"}. The abstract method
\lstinline{alg} in \lstinline{Refactor} can be any algebra of \lstinline{SubAlg}, simply for delegation. By passing a concrete implementation
to \lstinline{alg}, \lstinline{Refactor} will be able to apply transformation to an \lstinline{ExprAlg} expression before \lstinline{alg} returns the result.

\subsection{Parsing with Object Algebras}\label{subsec:parsingwithoa}

Just as shown above, \lstinline{ExprAlg} is defined for a small language in an Object-Algebra style. Furthermore,
A parser can be defined for this whole language at once. Since Object Algebras represent ``objects'' implicitly as functions like \lstinline{ExprAlg[E] =>} \lstinline{E} for generic \lstinline{E}, such a parser should consume an algebra from its parameter, then return a value of \lstinline{Fix[Parser[E]]} as in open recursion, where \lstinline{E} is abstract.
\begin{lstlisting}
trait ExprParser[E] {
    val pE : ExprAlg[E] => Fix[Parser[E]] = alg => p =>
        ident ^^ alg.varE |||
        p ~ p ^^ { case e1 ~ e2 => alg.appE(e1, e2) }
}
\end{lstlisting}
It is observed that \lstinline{pE} is in fact a parser generator, where \lstinline{alg} appears as a parameter, which can be any algebra, and after \lstinline{alg} is fed, it returns a \lstinline{Fix[Parser[E]]}. Again \lstinline{p} is the explicit self-reference of the open parser. Inside the body, the algebra is invoked correspondingly for all cases right after parsing. Note that the two cases are combined using the original \lstinline{|||} operator. To make \lstinline{alg} as a generic algebra, the enclosing trait of the parser, namely \lstinline{ExprParser}, is abstracted over \lstinline{E}. Behaviors, or algebras that can be fed to such a parser, are expected to be defined independently from the Object Algebra interface, just like the algebra \lstinline{ExprPrint} in Section~\ref{subsec:objectalgebras}.

To explain the modularity, we again add literals to expressions, but instead of using inheritance as Section~\ref{subsec:objectalgebras}, \lstinline{LitAlg} is defined independently together with its own parser and algebra:
\begin{lstlisting}
trait LitAlg[E] {
    def litE(n : Int) : E
}

trait LitPrint {
    def litE(n : Int) = n.toString
}

trait LitParser[E] {
    val pE : LitAlg[E] => Fix[Parser[E]] = alg => p =>
        numericLit ^^ alg.litE
}
\end{lstlisting}
In this case, \lstinline{LitAlg} is yet another small language, which is self-contained. To merge the two languages together, we require their grammars
to be merged, and we obtain a combined language, whose parser integrates the two small parsers using alternative. In Scala we use compound types, namely use \lstinline{"with"} for two Object Algebra interface types, to avoid polluting the namespace.

\begin{lstlisting}
trait ExprLitParser[E] {
    val pE : (ExprAlg[E] with LitAlg[E]) => Fix[Parser[E]] = alg => p => {
        val pExprE = new ExprParser[E](){}.pE(alg)(p)
        val pLitE = new LitParser[E](){}.pE(alg)(p)
        pExprE ||| pLitE
    }
}
\end{lstlisting}

%But we can only abstract over the whole interface type, instead of working on its type parameters, as the number of type parameters is not fixed, where we will discuss more on multiple syntax later. In summary, \lstinline{"|||"} produces a combined parser, which takes a compound algebra as its parameter.

In client code, a user can either define a new function with type \lstinline{(ExprAlg[E] with LitAlg[E]) =>} \lstinline{Parser[E]} based on \lstinline{pE}, which implies an algebra is applied after \lstinline{fix}, or simply feed an algebra to \lstinline{pE} at first, then obtain the result from \lstinline{fix}:
\begin{lstlisting}
trait ExprLitPrint extends ExprPrint with LitPrint
val parsePrint = fix(new ExprLitParser[String](){}.pE(new ExprLitPrint(){}))
val result = runParser(parsePrint)("x 3") // prints "(x 3)"
\end{lstlisting}
\haoyuan{Is there a way to avoid defining ExprLitPrint?}

### 2016-11-15 paper/sections/OpenRecursion.tex
Some changes. Will reorganize some stuff in next commit.

abstract class Expr
case class Var(x : String) extends Expr
val pVar : Parser[Expr] = ident ^^ Var
\end{lstlisting}
Note that we are omitting lexing and demo code here. And we use \lstinline{Parser} as a type synonym for \lstinline{PackratParser} throughout the paper. Furthermore, we would like the parser to support applications.
Not only the implementation of that parser is needed, but also its corresponding datatype:
\begin{lstlisting}
case class App(e1 : Expr, e2 : Expr) extends Expr
val pApp : Parser[Expr] = pVar ~ pVar ^^ { case e1 ~ e2 => App(e1, e2) }
\end{lstlisting}
Such a parser \lstinline{pApp} can only parse one-layer applications like \lstinline{"x y"}. Hence we revise it using recursion:
\begin{lstlisting}
val pApp : Parser[Expr] = {
    val p = pVar ||| pApp
    p ~ p ^^ { case e1 ~ e2 => new App(e1, e2) }
}
\end{lstlisting}
Note that we are composing \lstinline{pVar} and \lstinline{pApp} using alternative. Alternation is a general operation to compose different parsers, since they represent different grammar rules with the logical relation ``or'' between one another. In this example, both sub-expressions of an application can be either a single variable or again an application. Now \lstinline{pApp} manages to parse an arbitrary number of applications like \lstinline{"x y z ..."} in a right-associative way.
Nevertheless, even though we are not considering operations but the creation of objects only, such an approach does not turn out to be a modular extension, because the grammar we want for an expression is actually:
\begin{lstlisting}
e ::= x | e ' ' e | ...
\end{lstlisting}
Note that such a recursive \lstinline{"e"} in the second case should not only include all the existing cases like \lstinline{pApp} and \lstinline{pVar}, but also take future extensions into account (the ellipsis we used above, for instance, \lstinline{e '+'} \lstinline{e}). This is where we are inspired to use open recursion. To define \lstinline{pApp} as an ``open'' function, we add a parameter \lstinline{p} to it, representing the explicit self-reference of our whole parser, open to future extensions. Note that \lstinline{p} should be defined as a by-name parameter using \lstinline{"=>"}.
\begin{lstlisting}
val pApp : (=> Parser[Expr]) => Parser[Expr] =
    p => p ~ p ^^ { case e1 ~ e2 => new App(e1, e2) }
\end{lstlisting}
Here the type \lstinline{"(=>} \lstinline{Parser[Expr])} \lstinline{=>} \lstinline{Parser[Expr]"} is exactly what we want for a parser. Similarly \lstinline{pVar} is redefined also as a function using the explicit fix-point \lstinline{p}. To be more convenient for composition, we define an auxiliary combinator \lstinline{"|||"} for parsers of type \lstinline{"(=>} \lstinline{R)} \lstinline{=>} \lstinline{Parser[E]"} with the same \lstinline{R} and \lstinline{E}. Such a function is overloaded and can be invoked as an infix operator when defined in an implicit class. It is really implemented using the general \lstinline{merge} function in Section~\ref{subsec:openrecursion}.
 \begin{lstlisting}
val pVar : (=> Parser[Expr]) => Parser[Expr] = p => ident ^^ { x => new Var(x) }

implicit class Combinator[E, R](x : (=> R) => Parser[E]) {
    type H[A, B] = Parser[E]
    def |||(y : (=> R) => Parser[E]) : (=> R) => Parser[E] = {
        val alt : Parser[E] => Parser[E] => Parser[E] = a => b => a ||| b
        merge[R, Parser[E], Parser[E], H](alt, x, y)
    }
}
\end{lstlisting}
Now since \lstinline{pVar} and \lstinline{pApp} are defined independently, we are able to compose them using \lstinline{|||}:
\begin{lstlisting}
val pVarApp : (=> Parser[Expr]) => Parser[Expr] = pVar ||| pApp
\end{lstlisting}
Here \lstinline{pVarApp} is the combination of parsers. But what we really want is something that has type \lstinline{Parser[Expr]}. One can be soon reminded that in Section~\ref{subsec:openrecursion}, we use \lstinline{fix} to get the fix-point of a function, in this case we write \lstinline{fix(pVarApp)}, which manages to parse an arbitrary number of applications.

@@

But our story does not tend to stop here, as we have observed some limitations of the current approach. Specifically, one is that extensibility of data operations is not guaranteed. Suppose we want new operations like pretty-printing on expressions, we may 1) add a method \lstinline{print} to all case classes with their own interpretations; 2) define a single method \lstinline{print} and do pattern-matching on case classes inside. When adding new data variants and operations, both require existing code to be modified, thus lead to the lack of code reuse and modularity.

>>>>

abstract class Expr
case class Var(x : String) extends Expr
val pVar : Parser[Expr] = ident ^^ Var
\end{lstlisting}
Note that we are omitting lexing and demo code here. And we use \lstinline{Parser} as a type synonym for \lstinline{PackratParser} throughout the paper. Furthermore, we would like the parser to support applications.
Not only the implementation of that parser is needed, but also its corresponding datatype:
\begin{lstlisting}
case class App(e1 : Expr, e2 : Expr) extends Expr
val pApp : Parser[Expr] = pVar ~ pVar ^^ { case e1 ~ e2 => App(e1, e2) }
\end{lstlisting}
Such a parser \lstinline{pApp} can only parse one-layer applications like \lstinline{"x y"}. Hence we revise it using recursion:
\begin{lstlisting}
val pApp : Parser[Expr] = {
    val p = pVar ||| pApp
    p ~ p ^^ { case e1 ~ e2 => App(e1, e2) }
}
\end{lstlisting}
Note that we are composing \lstinline{pVar} and \lstinline{pApp} using alternative. Alternation is a general operation to compose different parsers, since they represent different grammar rules with the logical relation ``or'' between one another. In this example, both sub-expressions of an application can be either a single variable or again an application. Now \lstinline{pApp} manages to parse an arbitrary number of applications like \lstinline{"x y z ..."} in a right-associative way.
Nevertheless, even though we are not considering operations but the creation of objects only, such an approach does not turn out to be a modular extension, because if numeric literals are further accepted in our small language, the grammar becomes:
\begin{lstlisting}
e ::= x | e ' ' e | n
\end{lstlisting}
Besides writing a new parser for literals, we need to rewrite the existing \lstinline{pApp}, since applications can now contain numbers, like \lstinline{"x 3"} and so on.
\begin{lstlisting}
case class Lit(n : Int) extends Expr
val pLit : Parser[Expr] = numericLit ^^ Lit
val pApp : Parser[Expr] = {
    val p = pVar ||| pApp ||| pLit
    p ~ p ^^ { case e1 ~ e2 => App(e1, e2) }
}
\end{lstlisting}
We have observed that \lstinline{pApp} is indeed constructed in a non-modular way, because the \lstinline{p} standing for recursive parser needs to be updated on every extension. Actually in the grammar, the recursive \lstinline{"e"} in the second case should not only include all the existing cases like \lstinline{pApp} and \lstinline{pVar}, but also take future extensions into account (we can have arbitrarily more cases). This is where we are inspired to use open recursion. To define \lstinline{pApp} as an ``open'' function, we add a parameter \lstinline{p} to it, representing the explicit self-reference of our whole parser, open to future extensions. Note that \lstinline{p} should be defined as a by-name parameter using \lstinline{"=>"}.
\begin{lstlisting}
val pApp : Fix[Parser[Expr]] = p => p ~ p ^^ { case e1 ~ e2 => App(e1, e2) }
\end{lstlisting}
Here the type \lstinline{Fix[Parser[Expr]]}, namely \lstinline{(=>} \lstinline{Parser[Expr])} \lstinline{=>} \lstinline{Parser[Expr]} is exactly what we want for a parser. Similarly \lstinline{pVar} is redefined also as a function using the explicit fix-point \lstinline{p}. To be more convenient for composition, we define an auxiliary combinator \lstinline{"|||"} for parsers of type \lstinline{(=>} \lstinline{R)} \lstinline{=>} \lstinline{Parser[E]} with the same \lstinline{R} and \lstinline{E}. Such a function is overloaded and can be invoked as an infix operator when defined in an implicit class. It is really implemented using the general \lstinline{merge} function in Section~\ref{subsec:openrecursion}.
\begin{lstlisting}
val pVar : Fix[Parser[Expr]] = p => ident ^^ Var

implicit class Combinator[E, R](x : (=> R) => Parser[E]) {
    type H[A, B] = Parser[E]
    def |||(y : (=> R) => Parser[E]) : (=> R) => Parser[E] = {
        val alt : Parser[E] => Parser[E] => Parser[E] = a => b => a ||| b
        merge[R, Parser[E], Parser[E], H](alt, x, y)
    }
}
\end{lstlisting}
Now since \lstinline{pVar} and \lstinline{pApp} are defined independently, we are able to compose them using \lstinline{|||}:
\begin{lstlisting}
val pVarApp : Fix[Parser[Expr]] = pVar ||| pApp
\end{lstlisting}
Here \lstinline{pVarApp} is the combination of two parsers. But what we really want is something that has type \lstinline{Parser[Expr]}. One can be soon reminded that in Section~\ref{subsec:openrecursion}, we use \lstinline{fix} to get the fix-point of a function, in this case we write \lstinline{fix(pVarApp)}, which manages to parse an arbitrary number of applications.

@@

But our story does not tend to stop here, as we have observed some limitations of the current approach. Specifically, one is that extensibility of data operations is not guaranteed, though we obtain the extensibility of data variants by defining an arbitrarily large number of new case classes. Suppose we want new operations like pretty-printing on expressions, we may 1) add a method \lstinline{print} to all case classes with their own interpretations; 2) define a single method \lstinline{print} and do pattern-matching on case classes inside. When adding new data variants and operations, both require existing code to be modified, thus affect code reuse and modularity.
